### 2023-10-23 design/gru.tex
Bunch of changes, added design gru

\section{\ac{LSTM} analog implementation}\label{sec:lstmCircuit}

@@

This section describes the circuit of an \ac{LSTM} with an input vector of size $n_i$, a $n_h$ hidden states, a serial size of $n_s$ and $n_{ts}$ time steps. $n_o=n_h/n_s$ is going to be used for future references in this section. In order for the crossbar array to be used $n_o$ must be an integer, in other words, $n_s$ must divide $n_h$. The circuit (shown in \cref{circt:lstm}) is pretty complex and contains numerous parts that require explaination.

@@

The system is built using crossbar array (\cref{sec:xbarCircuit}) with ($n_i+n_h+1$,$n_o$, $n_s$) as parameters.

@@

The wires coming into the crossbar are a bus of size $n_i+n_h+1$ and the output of the crossbar is a bus of size $n_o$ (\cref{sym:xbar}). This is why everything in the system apart from the crossbar arrays is only shown once in \cref{circt:lstm} but in reality those components are present $n_o$ times. Those extra components are needed in order for the parallel channels to work.

\subsection{Doubling memory cells}

\subsubsection{Feedback hidden states}

In order for the system to work in serial mode, the memory cells of the output need to be doubled. This is done in \cref{circt:lstm}, and allows for the hidden states to be saved for the next stage. Indeed, if using the system in serial mode with only one line of memory cells, the old hidden state ($\overrightarrow{h_{t-1}}$) is slowly overwritten by the current hidden state ($\overrightarrow{h_t}$). As soon as the first $n_o$ serial values are computed, they are going to override the old ones, still required for the following serial values. \Cref{tim:lstmMemcell} shows that the value in the memory cell changes too early in the cycle and gives a bad input for the next serial values. The value given to the input of the \ac{LSTM} should always be the old hidden state ($\overrightarrow{h_{t-1}}$).

\begin{figure}[H]
  \centering
  \begin{tikztimingtable}
    $e_{n_s-1}$ & 3H8Lx \\
    $e_0$ & 3L4H4Lx \\
    $e_1$ & 7L4Hx \\
    $m_{i,0}$ & 3D{$h_{t-1,i,0}$}8D{$h_{t,i,0}$}x \\
    $m_{i,1}$ & 7D{$h_{t-1,i,1}$}4D{$h_{t,i,1}$}x \\
    \extracode
    \tablerules
    %\draw (0,0) circle (2pt); % Origin
    \begin{pgfonlayer}{background}
      \vertlines[help lines]{3.05}
      %\vertlines[red]{1.6,5.6,15.6}
      %\vertlines[blue]{3.6,9.6,15.6}
    \end{pgfonlayer}
  \end{tikztimingtable}
  \caption{Time diagram of the values in the different memory cells ($m_{i,j}$ being the value stored in the $i^{th}$ parallel memory cells for the $j^{th}$ serial value). It is assumed that $n_s>1$ (the system is in serial mode). This graph does not take into account the pauses $e_{next}$ between the time steps.}
  \label{tim:lstmMemcell}
\end{figure}

Using two memory cells when in serial mode allows for this issue to be elevated. The values are transfered from one level to an other at the end of the time step. That way, the values are updated safely during each time step, without changing the final output.

When the system is running is a fully parallel mode, doubling the memory cells still causes a problem. The enable in and enable out flags can't be high at the same time when the memory cell has a feedback connection, the signal would interfere with it self and change its own value. Potential solutions are discussed in \cref{subsec:noDoubleMemcell}.

\subsubsection{Cell states}

Cell states have the same solution to a similar problem. Here the problem is that the memory cell has a feedback connection and needs to be activated within one serial step. The value of the old cell state needs to be used to compute the current values for the hidden states ($\overrightarrow{h_t}$) and cell states ($\overrightarrow{c_t}$). The issue is that at the value in the memory cell needs to be conserved until the next serial step.  (\cref{subsec:noDoubleMemcell}).

Once again, when the system is running in fully parallel mode, only one memory cell line is required for a normal behavior.

\subsection{Serialization/Parallelization}
\label{subsec:serpar}

Using an \ac{LSTM} in serial mode is very beneficial as it divides the number of point wise circuit by $n_s$.

There is a sweet spot for the serial size ($n_s$) :

\begin{itemize}
  \item When the serial size ($n_s)$ increases, the inference time increases with a factor of $O(n_s)$ while the onChip area decrease with the same factor.
  \item When the serial size ($n_s)$ decreases, similarly, the onChip area increases with a factor of $O(n_s)$ while the tinference time decreases by the same factor.
\end{itemize}

The overall onChip area is linked to the number of hidden states ($n_h$) by $O(n_h^2)$. This means that the serial size ($n_s$) will be set depending on the limiting factor of our system (onChip area or inference time).

@@

The symbol of the \ac{LSTM} circuit can be found in \cref{sym:lstm}. It depends on several parameters, the number of inputs ($n_i$), the number of hidden states ($n_h$), the serial size of the system ($n_s$) and the number of time steps ($n_{ts}$).

\begin{figure}[H]
  \centering
  \includesvg[height=2.5cm]{lstm/lstmSymbol}
  \caption{Symbol used for the \ac{LSTM} circuit}
  \label{sym:lstm}
\end{figure}

@@

\begin{equation}
  \begin{array}{3}
    A_{lstm}(n_i,n_h,n_s)=4\cdot A_{xbar}(n_i,\frac{n_h}{n_s},n_s)\\+ \frac{n_h}{n_s}\cdot (5\cdot A_{af}+3\cdot A_{voltMult}+3\cdot A_{R_{amp0}}+2\cdot A_{R_{amp1}}+2\cdot A_{opAmp})\\+5\cdot n_h\cdot A_{memcell}
  \end{array}
\end{equation}

With $A_{CMOS}$ being the same than for the memory cell (\cref{sec:memcell}), $A_{CMOS}=1.2\cdot 10^{-14}m^2$, and the feature size of typical memristor that can be fabricated at INESC is $3\mu m$, which would make the approximate area for a memristor $A_{memristor}=9\mu m^2=9\cdot 10^{-12} m^2$.%TODO check info and inesc citation

>>>>

\section{\acs{GRU} analog implementation}\label{sec:gruCircuit}

@@

This section describes the circuit of an encoder \ac{GRU} with an input vector of size $n_i$, a $n_h$ hidden states and $n_{ts}$ time steps. The decoder \ac{GRU} could also be implemented in an analog circuit, but the choice was made to focus on the encoder \ac{GRU}. The \ac{GRU} is by its nature very similar to the \ac{LSTM}.

@@

The system is built, once again, using crossbar array (\cref{sec:xbarCircuit}) with ($n_i+n_h+1$,$n_o$, $n_s$) as parameters.

@@

The wires coming into the crossbar are a bus of size $n_i+n_h+1$ and the output of the crossbar is a bus of size $n_h$ (\cref{sym:xbar}). This is why everything in the system apart from the crossbar arrays is only shown once in \cref{circt:lstm} but in reality those components are present $n_o$ times. Those extra components are needed in order for the parallel channels to work.

\subsection{Serialization/Parallelization}
\label{subsec:gruSerPar}

As opposed to the to the \ac{LSTM}, the \ac{GRU} can not be serialized. This is due to a mathematical issue.

Indeed in the equation of the candidate hidden state is a problem for the serialization of the circuit. The \cref{eq:candActivG} from \cref{sec:gru} is repeated in \cref{eq:serProb} for better lisibility of the thesis.

\begin{equation}\label{eq:serProb}
  \overrightarrow{\hat{h_t}}=tanh(\overrightarrow{x_t}\cdot w_{hx}+(\overrightarrow{r_t}\odot\overrightarrow{h_{t-1}}) \cdot w_{hh} + \overrightarrow{b_h})
\end{equation}

As \cref{eq:serProb} shows, the reset gate's results vector needs to be multiplied with the the hidden state's previous vector. However, while the hidden state's previous vector is fully available in the memory cells, the reset gate's vector needs to be computed using the current inputs.

If the circuit were serialized, a full cycle would be required to compute the reset gate, that is itself required to compute the candidate hidden state.

That is why, the \ac{GRU} would take twice as much time to compute a time step if it were serialized. It is not worth it to double the computation time to save a bit of onChip area.

@@

The symbol of the \ac{GRU} circuit is available in \cref{sym:lstm}. It depends on a few parameters namely, the number of inputs ($n_i$), the number of hidden states ($n_h$) and the number of time steps ($n_{ts}$).

\begin{figure}[H]
  \centering
  \includesvg[height=2.5cm]{gru/gruSymbol}
  \caption{Symbol used for the \ac{GRU} circuit}
  \label{sym:lstm}
\end{figure}

@@

\begin{equation}
  \begin{array}{3}
    A_{gru}(n_i,n_h)=4\cdot A_{xbar}(n_i,\frac{n_h}{n_s},n_s)\\+ \frac{n_h}{n_s}\cdot (5\cdot A_{af}+3\cdot A_{voltMult}+3\cdot A_{R_{amp0}}+2\cdot A_{R_{amp1}}+2\cdot A_{opAmp})\\+5\cdot n_h\cdot A_{memcell}
  \end{array}
\end{equation}
